# Min. Dist Access Analysis {#min-distance}

## Overview {#CA-research-question}

Spatial Access to specific resource is often considered a multidimensional concept where accessibility can be measured on affordability, availability, accommodation & acceptability [**include reference**]. Distance to the nearest resource is a common metric used to capture the availability of a resource, and in this tutorial we demonstrate how to calculate a minimum distance value from a zip code centroid to a set of resources, such as locations of methadone clinics. Each zip code will be assigned a "minimum distance access metric" as a value that indicates access to resources from that zip code. Our objectives are thus to:

* Generate centroids from areal data
* Calculate minimum distance from resources to area centroids
* Overlay resources and new minimum distance metric

## Environment Setup {#CA-setup}
To replicate the codes & functions illustrated in this tutorial, you’ll need to have R and RStudio downloaded and installed on your system. This tutorial assumes some familiarity with the R programming language.

### Packages used {#CA-packages1}
We will use the following packages in this tutorial:

- `sf`: to manipulate spatial data
- `tmap`: to visualize and create maps
- `units`: to convert units within spatial data

### Required Inputs and Expected Outputs {#CA-i-o}
Our inputs will be:

- a CSV file with the locations of our resources ("chicago_methadone.csv"), and 
- a zip code boundary file ("chicago_zips.shp"). 

We will calculate the minimum distance between the resources and the centroids of the zip codes, then save the results as a shapefile and as a CSV. Our final result will be a shapefile/CSV with the minimum distance value for each zip. 

### Load the packages {#CA-packages2}
Load the libraries for use.

```{r load}
library(sf)
library(tmap)
library(units)
```

## Load data {#CA-read-resources}
First, load in the MOUD resources shapefile. Let's take a look at the first few rows of the dataset.

```{r}
meth_sf <- st_read("methadoneClinics.shp")
head(meth_sf)
```

If you don't have a shapefile of your data, but already have geographic coordinates as two columns in your CSV file, you can still use this tutorial. A reminder of how to transform your CSV with coordinates into a spatial data frame in R can be found [here](link to end of the geocoding tutorial).

Next, we'll load Chicago zip code boundaries.

```{r read zips}
chicago_zips <- read_sf("data/chicago_zips.shp")
```

We can quickly plot our data for to confirm they loaded correctly, here using an interactive map:
```{r}
tm_shape(chicago_zips) +
  tm_borders() +
tm_shape(meth_sf) +
  tm_dots(col = "blue", size = 0.2)
```

## Calculate centroids  {#CA-calculate-centroids}

Now, we will calculate the centroids of the zip code boundaries. We will first need to project our data, which means change it from latitude and longitude to meaningful units, like ft or meters, so we can calculate distance properly. We'll use the Illinois State Plane projection, with an EPSG code of 3435.

**Aside**: To find the most appropriate projection for your data, do a Google Search for which projection works well - for state level data, each state has a State Plane projection with a specific code, known as the EPSG. I use epsg.io to check projections - here's the [New York State Plane](https://epsg.io/2263) page.

Use the `st_transform` function to change the projection of the data. Notice how the values in `geometry` go from being relatively small (unprojected, lat/long) to very large (projected, in US feet).

```{r project-zips}
chicago_zips <- st_transform(chicago_zips, 3435)

chicago_zips
```

Then, we will calculate the centroids:

```{r calc-centroids}
chicago_centroids <- st_centroid(chicago_zips)

chicago_centroids
```

For each zip code, this will calculate the centroid, and the output will be a point dataset.

Plot to double check that everything is ok. The `st_geometry()` function will once again just return the outline:

```{r plot-zips-centroids}
plot(st_geometry(chicago_zips))
plot(st_geometry(chicago_centroids), add = TRUE, col = "red")
```

### Visualize & Confirm 

Once again, we can create an interactive map:

```{r plot-interactive-zips-centroids}
tm_shape(chicago_zips) +
  tm_borders() +
tm_shape(chicago_centroids) +
  tm_dots()
```


## Standardize CRS {#CA-correct-projection}

If we immediately try to calculate the distance between the zip centroids and the locations of the resources using the `st_distance` function, we'll get an error:

```{r eval=FALSE}
st_distance(chicago_centroids, meth_sf, by_element = TRUE)
```

```
Error in st_distance(chicago_centroids, meth_sf, by_element = TRUE) : st_crs(x) == st_crs(y) is not TRUE
```
Why is there an error? Because the projection of the centroids and the resource locations don't match up. Let's project the resource locations so that they match the projection of the centroids.

First, use the `st_crs` function to check that the coordinate reference system (or projection) is the same. They're not, so we have to fix it.

```{r check-crs-1}
st_crs(chicago_centroids)
st_crs(meth_sf)
```

We'll take the CRS from the zip code centroids data, and use it as input to `st_transform` applied to the methadone clinics data.

```{r project resources}
new_crs <- st_crs(chicago_centroids)
new_crs

meth_sf <- st_transform(meth_sf, new_crs)
```

If we check the CRS again, we now see that they match. **Mismatched projections are a commonly made mistake in geospatial data processing.**

```{r check-crs-2}
st_crs(chicago_centroids)
st_crs(meth_sf)
```

Now we have the zip boundaries, the centroids of the zips, and the resource locations, as shown below. Next, we will calculate the distance to the nearest resource from each zip code centroid.

```{r plot-zips-centroids-resources}
plot(st_geometry(chicago_zips))
plot(st_geometry(chicago_centroids), col = "red", add = TRUE)
plot(st_geometry(meth_sf), col = "blue", add = TRUE)
```

## Calculate min distance {-#CA-distance}

First, we'll identify the resource that is the closest to a zip centroid using the `st_nearest_feature` function. (It will return the index of the object that is nearest, so we will subset the resources by the index to get the nearest object.)

```{r find nearest resource}
nearest_clinic_indexes <- st_nearest_feature(chicago_centroids, meth_sf)

nearest_clinic <- meth_sf[nearest_clinic_indexes,]

nearest_clinic
```

Then, we will calculate the distance between the nearest resource and the zip code centroid with the `st_distance` function. As shown above, make sure both of your datasets are projected, and in the same projection, before you run `st_distance`.

```{r calc min dists}
min_dists <- st_distance(chicago_centroids, nearest_clinic, by_element = TRUE)

min_dists
```

This is in US feet. To change to a more meaningful unit, such as miles, we can use the `set_units()` function:

```{r convert units}
min_dists_mi <- set_units(min_dists, "mi")

min_dists_mi
```

### Merge data back to Area Data 
We then rejoin the minimum distances to the zip code data, by column binding `min_dists_mi` to the original `chicago_zips` data.

```{r rejoin}
min_dist_sf <- cbind(chicago_zips, min_dists_mi)
min_dist_sf
```

### Visualize & Confirm
We can now visualize the zip-level access to methadone clinics using our new access metric, using the `tmap` package.

```{r}
tmap_mode("plot")

tm_shape(min_dist_sf) +
  tm_polygons("min_dists_mi", 
              title = "Minimum Distance (mi)") +
  tm_layout(main.title = "Minimum Distance from Zip Centroid\n to Methadone Clinic",
            main.title.position = "center",
            main.title.size = 1)
```

Access by zip code can also be combined with locations of resources:

```{r}
tm_shape(min_dist_sf) +
  tm_polygons("min_dists_mi", style = 'jenks',
              title = "Minimum Distance (mi)") +
  tm_shape(meth_sf) +
  tm_dots(size = 0.2) +
  tm_layout(main.title = "Minimum Distance from Zip Centroid\n to Methadone Clinic",
            main.title.position = "center",
            main.title.size = 1)
```


```{r}
head(chicago_zips)
```

## Save Data {#CA-save}

To save our final result to a CSV:

```{r, eval=FALSE, write-csv}
write_sf(min_dist_sf, "chizips_Access.csv")
```

We can also write out this data to a shapefile format:

```{r, eval=FALSE, write-shp}
write_sf(min_dist_sf, "chizips_Access.shp")
```
