# Link Community Data {}

## Overview 

Stuff

* items

## Environment Setup 
To replicate the codes & functions illustrated in this tutorial, you’ll need to have R and RStudio downloaded and installed on your system. This tutorial assumes some familiarity with the R programming language.

### Packages used 
We will use the following packages in this tutorial:

- `sf`: to manipulate spatial data
- `tmap`: to visualize and create maps
- `units`: to convert units within spatial data

### Required Inputs and Expected Outputs  
Our inputs will be:

- a SHP zip code boundary file ("chicago_zips.shp"), a
- a CSV file with COVID case data from the city data portal, and 
- a SHP file with the locations of our resources ("methadoneClinics.shp"),  

We will calculate the minimum distance between the resources and the centroids of the zip codes, then save the results as a shapefile and as a CSV. Our final result will be a shapefile/CSV with the minimum distance value for each zip. 

### Load the packages 
Load the libraries for use.

```{r load}
library(sf)
library(tmap)
```

## Load data 
First we'll load the update zip code dataset from a previous labs.

```{r}
chicago_zips <- read_sf("data/chicago_zips.shp")
str(chicago_zips)
```
```{r}
meth_sf <- st_read("methadoneClinics.shp")
```

Next, we'll load some new data we're interested in joining in: Chicago COVID-19 Cases, Tests, and Deaths by ZIP Code, found on the city data portal [here](https://data.cityofchicago.org/Health-Human-Services/COVID-19-Cases-Tests-and-Deaths-by-ZIP-Code/yhhz-zm2v). We'll load in a CSV and inspect the data:

```{r}
COVID <- read.csv("data/COVID-19_Cases__Tests__and_Deaths_by_ZIP_Code.csv")
str(COVID)
```

## Clean & Merge Data

```{r}
#COVID1 <- COVID[, c("`ZIP Code`", "`Percent Tested Positive - Cumulative`","Population")]
#head(COVID1)
head(chicago_zips)
```

```{r}
COVID$GEOID10<- as.character(COVID$ZIP.Code)
```

Let's merge the data using the zip code geographic identifier, "ZIP Code" field, to bring in the the Percent Tested Positive - Cumalative dataset.

```{r}
new <- merge(chicago_zips, COVID, by = "GEOID10")
head(new)

```


```{r}
new$COVIDCaseRt <- new$Case.Rate...Cumulative 
str(new)

union.buffers<- st_read("data/mclinicarea.shp")

```


## Visualize Data

```{r}
tm_shape(new) +
  tm_polygons("COVIDCaseRt", style="jenks", pal="BuPu",
              title = "COVID Case Rate") +
  tm_shape(union.buffers) + tm_borders(col = "blue") +
  tm_shape(meth_sf) + tm_dots(col = "black", size = 0.2) 
```


